% !TEX root = % !TEX root = /Users/nai1ka/Projects/PerfX/Thesis/thesis.tex
\begin{abstract}
% skip one line to make the abstract start with indent


Mobile applications are updated frequently, and a new release can quietly degrade performance on real user devices, an effect known as a performance regression. Detecting such regressions is difficult because Android devices vary widely in hardware and operating system version, and because existing tools address only part of the problem. Pre-release profilers run under controlled conditions and add too much overhead for continuous use, while production monitoring tools either provide no automated alerting or rely on fixed absolute thresholds that must be set by hand for each metric and rarely fit every device. Release-to-release regression detection, in turn, has been studied mainly for homogeneous server software rather than for fragmented Android fleets. This thesis develops a system that combines continuous low-overhead post-release monitoring on Android with automated server-side regression detection. An embeddable Kotlin SDK collects user-perceived and resource metrics on real devices and uploads them to a Ktor backend in batches, where they are stored in ClickHouse and PostgreSQL and grouped by device performance cohort. A detection pipeline compares each release against the previous one using the relative shift of the 95th percentile, grouped by metric, screen, and cohort, and reports a regression when the shift exceeds a single relative threshold. This removes the need to calibrate absolute per-metric limits and attributes every alert to a specific release. The system was evaluated for runtime overhead and for detection accuracy. Enabling the SDK added at most 5.2 percentage points of CPU usage, 5.4 MB of memory, and 75 ms of cold-start time across the tested devices. On synthetic regressions spanning three types, three intensities, and three device cohorts, the pipeline detected 27 of 30 cases (90 percent) at the default 15 percent threshold, with one false alert in 15 control runs. The results show that continuous monitoring and automated, cohort-aware regression detection can be delivered together on real Android devices at a cost that users are unlikely to notice.
\end{abstract}