% !TEX root = % !TEX root = /Users/nai1ka/Projects/PerfX/Thesis/thesis.tex
\chapter{Conclusion}
\label{chap:conclusion}

This thesis addressed performance monitoring and regression detection for Android applications after release. The problem it tackled is that existing tools either provide no automated alerting or rely on fixed thresholds that have to be set and maintained by hand for every metric. On a fragmented device base, such absolute thresholds produce frequent false or missed alerts, and a slowdown introduced by a new release can go unnoticed as long as it stays within the configured limit.

To address this, a complete system was designed and implemented. A mobile SDK collects performance metrics on real user devices with low overhead and uploads them to the backend in batches. A Ktor backend stores the data in ClickHouse and PostgreSQL and assigns a device cohort to each record. A detection pipeline then compares consecutive releases by the relative P95 shift, grouped by metric, screen, and cohort, and notifies developers when a new release degrades performance. The system was evaluated for runtime overhead and for the accuracy of its regression detection. The results, together with the answers to the research questions, are reported in Chapter~\ref{chap:eval}.

The usefulness of the work lies in combining continuous post-release monitoring on real Android devices with automated, server-side detection of regressions between releases, a combination not offered by existing tools. Because detection is based on the relative change between two releases rather than on absolute limits, developers do not need to set or tune thresholds by hand. The comparison is also stratified by device cohort, so slower low-end hardware is not mistaken for a regression. The system is not intended to replace pre-release profiling tools. Instead, it complements them by covering the post-release period on real user devices.

\section{Future Work}
\label{sec:concl_future}

The most important next step is a real-world deployment on a live application with real users, which would test the system against naturally occurring regressions under varied hardware, usage patterns, and network conditions. The cohort classification could also be improved. Instead of relying on static hardware attributes such as RAM and CPU core count, cohorts could be derived from each device's observed performance, for example by grouping devices according to a lightweight on-device benchmark. Finally, the set of collected metrics could be extended with battery and network usage, which are also relevant to mobile user experience~\cite{Saraiva} but are not currently collected.
